Capstone Movielens Report

### 2025-02-08 reports/capstone-movielens-report.Rmd
`edx` dataset: analysis in progress...

# Loading the required libraries
library(caret)
library(cowplot)
library(data.table)
library(dplyr)
library(ggplot2)
library(ggthemes)
library(lubridate)
library(Metrics)
library(recosystem)
library(scales)
library(stringr)
library(tibble)
library(tidyr)
library(pak)

@@

The goal of the project is to build a Recommendation System using a [10M version of the MovieLens dataset](http://grouplens.org/datasets/movielens/10m/).
Following the [Netflix Grand Prize Contest](https://archive.nytimes.com/bits.blogs.nytimes.com/2009/09/21/netflix-awards-1-million-prize-and-starts-a-new-contest/index.html) requirements, we will evaluate the _Root Mean Square Error_ (_RMSE_) score defined as:

$$
\mbox{RMSE} = \sqrt{\frac{1}{N} \sum_{i,j}^{N} (y_{i,j} - \hat{y}_{i,j})^2}
$$

@@

Let us look into the details of the `edx` dataset:
``` {r echo = TRUE}
str(edx)
```
Note that we have 9000055 rows and six columns in there:  
```{r }
dim(edx)
```

@@

Next, we have 10677 different movies 
```{r}
n_distinct(edx$movieId)
```
and 69878 different users in the dataset:
```{r}
n_distinct(edx$userId)
```

The following code computes movie rating summaries by popular genres like Drama, Comedy, Thriller, and Romance:
```{r}
#library(stringr)
genres = c("Drama", "Comedy", "Thriller", "Romance")
sapply(genres, function(g) {
  sum(str_detect(edx$genres, g))
})
```

@@

Further analysis of the `edx` dataset was inspired by the article: A Movie Recommender System Using R - BEST[@MRS-R-BEST], from which the code and explanatory notes below were cited.

>>>>

# Loading the required libraries
library(caret)
library(cowplot)
library(data.table)
library(dplyr)
library(ggplot2)
library(ggthemes)
library(lubridate)
library(Metrics)
library(recosystem)
library(scales)
library(stringr)
library(tibble)
library(tidyr)

library(rafalib)
library(pak)

@@

The goal of the project is to build a Recommendation System using a [10M version of the MovieLens dataset](http://grouplens.org/datasets/movielens/10m/).
Following the [Netflix Grand Prize Contest](https://archive.nytimes.com/bits.blogs.nytimes.com/2009/09/21/netflix-awards-1-million-prize-and-starts-a-new-contest/index.html) requirements, we will evaluate the _Root Mean Square Error_ (_RMSE_) score, which, as shown in [Section 23.2 Loss function](https://rafalab.dfci.harvard.edu/dsbook-part-2/highdim/regularization.html#sec-netflix-loss-function) of the _Course Textbook_, is defined as:
$$
\mbox{RMSE} = \sqrt{\frac{1}{N} \sum_{i,j}^{N} (y_{i,j} - \hat{y}_{i,j})^2}
$$

@@

Let us look into the details of the `edx` dataset:
``` {r echo = TRUE}
str(edx)
```
Note that we have 9000055 rows and six columns in there:  
```{r }
dim_edx <- dim(edx)
print(dim_edx)
```

@@

Next, we have 10677 different movies 
```{r}
n_movies <- n_distinct(edx$movieId)
print(n_movies)
```
and 69878 different users in the dataset:
```{r}
n_users <- n_distinct(edx$userId)
print(n_users)
```

Now, note the expressions below which confirm the fact explained in [Section 23.1.1 Movielens data](https://rafalab.dfci.harvard.edu/dsbook-part-2/highdim/regularization.html#movielens-data) of the _Course Textbook_[@IDS2]:
```{r}
max_possible_ratings <- n_movies*n_users
sprintf("Maximum possible ratings: %s", max_possible_ratings)
sprintf("Rows in `edx` dataset: %s", dim_edx[1])
sprintf("Not every movie was rated: %s", max_possible_ratings > dim_edx[1])

As also explained in that section, we can think of these data as a very large matrix, with users on the rows and movies on the columns, with many empty cells. Therefore, we can think of a recommendation system as filling in the `NA`s in the dataset for the movies that some or all the users do not rate. A sample from the `edx` data below illustrates this idea[@IDS2_23-1-1]: 
```{r}
keep <- edx |> 
  dplyr::count(movieId) |> 
  top_n(4, n) |> 
  pull(movieId)

tab <- edx |> 
  filter(movieId %in% keep) |> 
  filter(userId %in% c(13:20)) |> 
  select(userId, title, rating) |> 
  mutate(title = str_remove(title, ", The"),
         title = str_remove(title, ":.*")) |>
  pivot_wider(names_from = "title", values_from = "rating")

print(tab)
```

The following plot of the matrix for a random sample of 100 movies and 100 users with yellow indicating a user/movie combination for which we have a rating shows how _sparse_ the matrix is[@IDS2_23-1-1]:
```{r sparsity-of-movie-recs, echo=TRUE, fig.width=3, fig.height=3, out.width="40%"}
users <- sample(unique(edx$userId), 100)

rafalib::mypar()
edx|> 
  filter(userId %in% users) |> 
  select(userId, movieId, rating) |>
  mutate(rating = 1) |>
  pivot_wider(names_from = movieId, values_from = rating) |> 
  (\(mat) mat[, sample(ncol(mat), 100)])() |>
  as.matrix() |> 
  t() |>
  image(1:100, 1:100, z = _ , xlab = "Movies", ylab = "Users")
```

Further observations highlighted there that, as we can see from the distributions the author presented, some movies get rated more than others, and some users are more active than others in rating movies:
```{r movie-id-and-user-hists, echo=FALSE, fig.width=6, fig.height=3}
p1 <- edx |> 
  count(movieId) |> 
  ggplot(aes(n)) + 
  geom_histogram(bins = 30, color = "black") + 
  scale_x_log10() + 
  ggtitle("Movies")

p2 <- edx |> 
  count(userId) |> 
  ggplot(aes(n)) + 
  geom_histogram(bins = 30, color = "black") + 
  scale_x_log10() + 
  ggtitle("Users")

gridExtra::grid.arrange(p2, p1, ncol = 2)
```

Taking into consideration these observations, we came up with a decision to use the data from the dataset only for users who have provided at least 100 ratings[@IDS2_23-1-1].

#### Movie genres
\
The following code computes movie rating summaries by popular genres like Drama, Comedy, Thriller, and Romance:
```{r}
#library(stringr)
genres = c("Drama", "Comedy", "Thriller", "Romance")
sapply(genres, function(g) {
  sum(str_detect(edx$genres, g))
})
```

@@

Further analysis of the `edx` dataset have been also inspired by the article mentioned above[@MRS-R-BEST], from which the code and explanatory notes below were cited.
